---
title: "Metabolomics-Norm-DE"
author: "Alessio Tovaglieri"
date: "10/12/2018"
output: html_document
---
```{r,results=FALSE, include=FALSE}
rm(list=ls())

library(readr)
library(dplyr)
library(gplots)
library(RColorBrewer)
library(rmarkdown)
library(ggplot2)
library(data.table)
library("pheatmap")
library("DESeq2")
library(limma)
library(readxl)
library(pcaMethods)
library(metabolomics)
library("tibble")

#functions
convert_num_to_df <- function (num)
{names=rownames(num)
temp_num=as.data.table(num)
row.names(temp_num)=names
return(temp_num)}

```

Directories and files 


```{r setup, include=FALSE}
abundance=read_csv("./data/metabolites_abundance_raw.csv")
data_meta=read_csv("./data/metabolites_ID.csv",na = "")

temp_abundance=abundance

temp_abundance=column_to_rownames(temp_abundance,var = 'Sample')

#Using metabolomics package that does not feed on tydy datasets...
#LOG TRANSFORM (needed before normalization)
temp_log_abundance=NULL
#run LogTransform
temp_log_abundance=LogTransform(temp_abundance, base = exp(1),
saveoutput = FALSE, outputname = "log_abundance")
log_abundance=as_data_frame(temp_log_abundance$output)
log_abundance=add_column(log_abundance,Sample=temp_log_abundance$samples,.before = 1)
temp_log_abundance=NULL
temp_abundance=NULL


#NORMALISE
temp_log_abundance=NULL
#move column to rownames to feed it to normnalize function
temp_log_abundance=column_to_rownames(log_abundance,var = 'Sample')
#run Normalise
temp_norm_log_abundance=Normalise(temp_log_abundance, method = c("median"), saveoutput = FALSE)
norm_log_abundance=as_data_frame(temp_norm_log_abundance$output)
norm_log_abundance=add_column(norm_log_abundance,Sample=temp_norm_log_abundance$samples,.before = 1)
temp_norm_log_abundance=NULL
temp_log_abundance=NULL

#Create a transposed version of the table 
t_log_abundance=NULL
t_log_abundance=log_abundance %>% select(-Group) %>% tidyr::gather(MetaID,valname,-Sample) %>% tidyr::spread(Sample,valname)


```
#plot median before and after normalization
```{r meta_stock_norm, results=FALSE, include=TRUE}

# Check distributions of RAW and NORMALIZED samples using BOXPLOTS

#BOXPLOT OF RAW DATA
temp_t_log_abundance= t_log_abundance %>% remove_rownames() %>% column_to_rownames("MetaID")

boxplot(temp_t_log_abundance, xlab="", ylab="Log2 counts per million",las=2,par(mar=c(8,8,4,2)),staplewex = 1)
title("Boxplots of raw data")
temp_t_log_abundance=NULL

#BOXPLOT OF NORMALISED DATA
#Create a transposed version of the table for boxplot
t_norm_log_abundance=NULL
t_norm_log_abundance=norm_log_abundance %>% select(-Group) %>% tidyr::gather(MetaID,valname,-Sample) %>% tidyr::spread(Sample,valname)

temp_t_norm_log_abundance= t_norm_log_abundance %>% remove_rownames() %>% column_to_rownames("MetaID")
#save normalised data transposed version
write.csv(t_norm_log_abundance, file="./res/t_norm_log_abundance.csv")
boxplot(temp_t_norm_log_abundance, xlab="", ylab="Log2 counts per million",las=2,par(mar=c(8,8,4,2)),staplewex = 1)
title("Boxplots of normalized data")
temp_t_norm_log_abundance=NULL



```






#dendogram
```{r meta_stock_dendo, results=FALSE, include=TRUE}
#change back to rownames to fgeed to dendogram
temp_norm_log_abundance=NULL
temp_norm_log_abundance=norm_log_abundance
temp_norm_log_abundance=column_to_rownames(temp_norm_log_abundance,var='Sample')
Dendrogram(as.data.frame(temp_norm_log_abundance),distmethod = "manhattan", aggmethod = "ward", main = "Dendrogram all metab norm", cex = 0.4)
temp_norm_log_abundance=NULL
```


##CALCULATE AND PLOT PCA USING ALL METABOLITES after norm.

```{r meta_stock_pca, results=FALSE, include=TRUE}
#assign values to needed variables
log.ir <- norm_log_abundance[,-c(1,2)]
ir.species <- norm_log_abundance[, 2]

ir.pca <- prcomp(log.ir,
                 center = TRUE,
                 scale. = TRUE) 

# extract PCA
PCA=ir.pca$x

PCA=convert_num_to_df(PCA)
#add back a group column
PCA$Group=norm_log_abundance$Group


#extract PCA summary
s=summary(ir.pca)
summary_variance=s$importance
summary_variance=convert_num_to_df(summary_variance)
#extract percent variation of the first 2 axis PC1 andPC2
percentVar <- round(100*as.numeric(summary_variance[2,1:2]))

p=ggplot(PCA, aes(PC1, PC2, color=Group, shape=Group)) +
  geom_point(size=5) +
  theme_classic() +
  ggtitle("PCA all metabolites (norm)") +
  theme(plot.title = element_text(hjust = 0.5),axis.text=element_text(size=10),legend.text = element_text(size=10),legend.title = element_text(size=10), axis.title=element_text(size=10) )+
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) 
p


# plot method
plot(ir.pca, type = "l")

#remove useless variables
log.ir=NULL
ir.species=NULL
ir.pca=NULL
s=NULL
summary_variance=NULL
percentVar=NULL
p=NULL
PCA=NULL
```

#calculate DE: Hmb vs Pre-ferm and Mmb vs Pre-ferm 

```{r meta_stock_de, results=FALSE, include=TRUE}

#define mydata
mydata=norm_log_abundance[,-c(1,2)]

#define design table
mydesign=NULL
#group H
mydesign$GroupH[norm_log_abundance$Group=="Hmb_stock_100"]=1
mydesign$GroupH[norm_log_abundance$Group!="Hmb_stock_100"]=0
#group M
mydesign$GroupM[norm_log_abundance$Group=="Mmb_stock_100"]=1
mydesign$GroupM[norm_log_abundance$Group!="Mmb_stock_100"]=0
#group P
mydesign$GroupP[norm_log_abundance$Group=="Pre-ferm_100"]=1
mydesign$GroupP[norm_log_abundance$Group!="Pre-ferm_100"]=0   
mydesign=as.data.table(mydesign)

#lmfit 
fit <- lmFit(t(mydata), mydesign)  
contrast.matrix=makeContrasts(GroupH - GroupP, GroupM - GroupP, GroupH - GroupM, levels = mydesign)

fit2 = contrasts.fit(fit, contrast.matrix)
fit2 = eBayes(fit2)
DE_H_vs_P = topTable(fit2, coef=1, adjust="fdr", lfc = log2(1.5), p.value = 0.05,number = 5073)
DE_M_vs_P = topTable(fit2, coef=2, adjust="fdr", lfc = log2(1.5), p.value = 0.05,number = 5073)


#convert to tibble data frame
DE_H_vs_P=as_data_frame(DE_H_vs_P)
DE_M_vs_P=as_data_frame(DE_M_vs_P)

#put back the row names as a column
DE_H_vs_P=rownames_to_column(DE_H_vs_P,var='MetabID')
DE_M_vs_P=rownames_to_column(DE_M_vs_P,var='MetabID')

#add 2 new columns with names and putative names
DE_H_vs_P=left_join(DE_H_vs_P,data_meta[,1:3],by = c("MetabID", "MetabID"))
DE_M_vs_P=left_join(DE_M_vs_P,data_meta[,1:3],by = c("MetabID", "MetabID"))


```

#define a variable with the union of the UPREGULATED metabolites in H vs P and M vs P

```{r meta_stock_upreg, results=FALSE, include=TRUE}


names_DE_H_vs_P_fcPOS=DE_H_vs_P[DE_H_vs_P$logFC>log2(1.5),]
names_DE_M_vs_P_fcPOS=DE_M_vs_P[DE_M_vs_P$logFC>log2(1.5),]

#do a union of the two names of list and keep only uniques names
names_H_M_vs_P_fcPOS=NULL
names_H_M_vs_P_fcPOS=c(names_DE_H_vs_P_fcPOS$MetabID,names_DE_M_vs_P_fcPOS$MetabID)
names_H_M_vs_P_fcPOS=unique(names_H_M_vs_P_fcPOS)
#remove variables
names_DE_H_vs_P_fcPOS=NULL
names_DE_M_vs_P_fcPOS=NULL
fit=NULL
fit2=NULL
mydata=NULL
mydesign=NULL
contrast.matrix=NULL
```

#feed the subset of norm_log_abundance (the one woth metabolites increased in H and M compared tor Pre) to the linearized model to calculate DE again

```{r meta_stock_de_subs, results=FALSE, include=TRUE}

mydata=norm_log_abundance[,names(norm_log_abundance) %in% names_H_M_vs_P_fcPOS]


#define design table

#group H
mydesign$GroupH[norm_log_abundance$Group=="Hmb_stock_100"]=1
mydesign$GroupH[norm_log_abundance$Group!="Hmb_stock_100"]=0
#group M
mydesign$GroupM[norm_log_abundance$Group=="Mmb_stock_100"]=1
mydesign$GroupM[norm_log_abundance$Group!="Mmb_stock_100"]=0
#group P
mydesign$GroupP[norm_log_abundance$Group=="Pre-ferm_100"]=1
mydesign$GroupP[norm_log_abundance$Group!="Pre-ferm_100"]=0   
mydesign=as.data.table(mydesign)

#lmfit 
fit <- lmFit(t(mydata), mydesign)  
contrast.matrix=makeContrasts(GroupH - GroupP, GroupM - GroupP, GroupH - GroupM, levels = mydesign)

fit2 = contrasts.fit(fit, contrast.matrix)
fit2 = eBayes(fit2)
DE_H_vs_M_RunOnPOSMetabOnly = topTable(fit2, coef=3, adjust="fdr", lfc = log2(1.5), p.value = 0.05,number = 5073)

#save the DE tables adding the metabolites name and put name info
#convert to tibble data frame
DE_H_vs_M_RunOnPOSMetabOnly=as_data_frame(DE_H_vs_M_RunOnPOSMetabOnly)
#put back the row names as a column
DE_H_vs_M_RunOnPOSMetabOnly=rownames_to_column(DE_H_vs_M_RunOnPOSMetabOnly,var='MetabID')
#add 2 new columns with names and putative names
DE_H_vs_M_RunOnPOSMetabOnly=left_join(DE_H_vs_M_RunOnPOSMetabOnly,data_meta[,1:3],by = c("MetabID", "MetabID"))
write.csv(DE_H_vs_M_RunOnPOSMetabOnly, file="./res/DE_H_vs_M.csv")


#remove variables
names_DE_H_vs_P_fcPOS=NULL
names_DE_M_vs_P_fcPOS=NULL
fit=NULL
mydata=NULL
mydesign=NULL
contrast.matrix=NULL
```



